Study sample characteristics in chronobiology and sleep research

### 2021-05-20 article.Rmd
updates

author: 
  - name          : "Selma Tir"
    affiliation   : "1,2"
    email         : "selma.tir@ndcn.ox.ac.uk"
    role:
      - Data curation
      - Investigation
      - Software
      - Visualization
      - Writing – original draft
      - Writing – review & editing
  - name          : "Rhiannon White"
    affiliation   : "1,3"
    email         : "r.white.5@warwick.ac.uk"
    role:         
      - Investigation
      - Writing – original draft
      - Writing – review & editing
  - name          : "Manuel Spitschan"
    affiliation   : "1,4,5"
    corresponding : yes    # Define only one corresponding author
    email         : "manuel.spitschan@psy.ox.ac.uk"
    role:
      - Conceptualization
      - Data curation
      - Funding acquisition
      - Investigation
      - Methodology
      - Project administration
      - Software
      - Writing – original draft
      - Writing – review & editing

@@

authornote: |
  Corresponding author: Manuel Spitschan (manuel.spitschan@psy.ox.ac.uk). This work ws funded by the Wellcome Trust (Sir Henry Wellcome Fellowship to MS 204686/Z/16/Z; Research Enrichment – Diversity & Inclusion WT 204686/Z/16/A) and Linacre College (Biomedical Sciences Junior Research Fellowship to MS).
abstract: |
  Many aspects of sleep and circadian physiology appear to be sensitive to participant-level characteristics. While recent research robustly highlights the importance of considering participant-level demographic information, it is not clear to what extent this information is available in the large body of already published literature. Here, we investigated study sample characteristics in the published sleep and chronobiology research over the past 40 years. 6,777 articles were identified and a random sample of 20% was included. The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored, and any reported aggregate summary statistics for these variables were recorded. We found that while >90% of studies reported age or sex, all other variables were reported in <10% of cases. Sex balance greatly changed over the years, from a ~3:1 male to female ratio in the 1990s to a near-equal representation in the 2010s. We found that the majority of studies report at least sex or age, while other variables are typically not reported. Reporting quality is highly variable, indicating an opportunity to standardize reporting guidelines for participant-level characteristics to facilitate meta analyses.

keywords          : "demographics, ethnicity, sex, research participants, reporting, publishing, meta-science"
wordcount         : "X"

bibliography      : references.bib

@@

@@

### Procedure

@@

### Article inclusion

@@

### Review and article extraction

@@

### Pre-registration

@@

### Materials, data and code availability

@@

### The rest  of the analysis will be conducted on included single studies
# Filter single studies
data <- data.included %>%
  filter(num_substudies == 1)
```

@@

### Number of analysed articles

@@

### Funding

@@

### Geographical location

@@

### Sample size

@@

### Age

@@

### Sex

@@

###  Ethnicity, education, profession and socio-economic status

@@

```{r, fig.cap="Reporting of education, ethnicity, profession and socio-economic status (SES).", warning=FALSE, message=FALSE}
figure10
```
  We furthermore examined the number of categories included for each of these demographic variables. In Figure 9, we show the number of categories reported for each variable among those articles that included it in a histogram. On overage, we find that the coding scheme for each of these variables in the included articles comprised between 3 and 4 variables.

@@

### Study focus

@@

### Analysis disaggregation

>>>>

author: 
  - name          : "Selma Tir"
    affiliation   : "1,2"
    email         : "selma.tir@ndcn.ox.ac.uk"
    role:
      - Data curation
      - Investigation
      - Software
      - Visualization
      - Writing – original draft
      - Writing – review & editing
  - name          : "Rhiannon White"
    affiliation   : "1,3"
    email         : "r.white.5@warwick.ac.uk"
    role:         
      - Investigation
      - Writing – original draft
      - Writing – review & editing
  - name          : "Manuel Spitschan"
    affiliation   : "1,4,5"
    corresponding : yes    # Define only one corresponding author
    email         : "manuel.spitschan@psy.ox.ac.uk"
    address       : Department of Experimental Psychology, University of Oxford, Anna Watts Building, Radcliffe Observatory Quarter, Woodstock Rd, Oxford, OX2 6GG, United Kingdom
    telephone     : +44 1865 281861
    role:
      - Conceptualization
      - Data curation
      - Funding acquisition
      - Investigation
      - Methodology
      - Project administration
      - Software
      - Writing – original draft
      - Writing – review & editing

@@

authornote: |
 This work was funded by the Wellcome Trust (Sir Henry Wellcome Fellowship to MS 204686/Z/16/Z; Research Enrichment – Diversity & Inclusion WT 204686/Z/16/A) and Linacre College (Biomedical Sciences Junior Research Fellowship to MS).
abstract: |
  Many aspects of sleep and circadian physiology appear to be sensitive to participant-level characteristics. While recent research robustly highlights the importance of considering participant-level demographic information, it is not clear to what extent this information is available in the large body of already published literature. Here, we investigated study sample characteristics in the published sleep and chronobiology research over the past 40 years. 6,777 articles were identified and a random sample of 20% was included. The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored, and any reported aggregate summary statistics for these variables were recorded. We found that while >90% of studies reported age or sex, all other variables were reported in <10% of cases. Sex balance greatly changed over the years, from a ~3:1 male to female ratio in the 1990s to a near-equal representation in the 2010s. We found that the majority of studies report at least sex or age, while other variables are typically not reported. Reporting quality is highly variable, indicating an opportunity to standardize reporting guidelines for participant-level characteristics to facilitate meta analyses.

keywords          : "demographics, ethnicity, sex, research participants, reporting, publishing, meta-science"
wordcount         : "3200"

bibliography      : references.bib
csl               : vancouver.csl

@@

# Summary

In this article, we addressed the question how representiative, diverse and inclusive published research articles in sleep and chronobiology are. We analysed a sample of >1,300 articles published in sleep research and chronobiology journals between 1979 and 2019 for its inclusion, reporting analysis of demographic variables of study participants, including age, sex, gender, ethnicity, level of education, socio-economic status, and professio.  We found that while >90% of studies reported age or sex, all other variables were reported in <10% of cases, with the frequency of reporting changing over time. We identify opportunities for improving the reporting of demographic variables.

# Research Agenda

Future research needs to identify:
1. Schemas for reporting demographic variables in a harmonised way across geographical and culutral contexts;
2. Gaps in the sleep and chronobiology literature with respect to understudied populations;
3. Understand the extent to which research practices run counter the inclusion of diverse populations in research in all stages of the research cycle, and how this can exacerbate health inequities.

# Practice Points

1. Published studies on circadian and sleep physiology should be carefully examined.
1. Reporting of demographic variables should be done carefully and systematically.
2. Inclusion and diversity of different populations across the field needs to be ensured.

@@

## Procedure

@@

## Article inclusion

@@

## Review and article extraction

@@

## Pre-registration

@@

## Materials, data and code availability

@@

## The rest  of the analysis will be conducted on included single studies
# Filter single studies
data <- data.included %>%
  filter(num_substudies == 1)
```

@@

## Number of analysed articles

@@

## Funding

@@

## Geographical location

@@

## Sample size

@@

## Age

@@

## Sex

@@

##  Ethnicity, education, profession and socio-economic status

@@

```{r, fig.cap="test", warning=FALSE, message=FALSE}
figure10
```
  We furthermore examined the number of categories included for each of these demographic variables. In Figure 9, we show the number of categories reported for each variable among those articles that included it in a histogram. On overage, we find that the coding scheme for each of these variables in the included articles comprised between 3 and 4 variables.

@@

## Study focus

@@

## Analysis disaggregation
